---
title: "Class 09: Candy Mini-Project"
author: "Zeyu Qi(A17342618)"
format: pdf
toc: true
---

## Background 

Today we are taking a small detor to analyze a dataset (that we have more intrinsic insight into) with the most useful analysis method we have learned thus far - Principal Component Analysis (PCA)

## Data Import
The data is all related to halloween candy and is from the 538 website. 

```{r}
candy_file <- "candy-data.csv"

candy = read.csv(candy_file, row.names=1)
head(candy)
```


> Q1. How many different candy types are in this dataset?

```{r}
nrow(candy)
```

> Q2. How many fruity candy types are in the dataset?

```{r}
sum(candy$fruity == 1)
```

> Q3. What is your favorite candy (other than Twix) in the dataset and what is it’s winpercent value?

```{r}
candy["Nerds", ]$winpercent
```

> Q4. What is the winpercent value for “Kit Kat”?

```{r}
candy["Kit Kat", ]$winpercent
```

> Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy["Tootsie Roll Snack Bars", ]$winpercent
```

There is a useful `skim()` function in the **skimr** package that can help give you a quick overview of a given dataset. Let's install this package and try it on our candy data. 
```{r}
# install.packages("skimr")
skimr::skim(candy)
```

> Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

Yes. Most columns are on a 0/1 binary scale. Sugarpercent, pricepercent, and winpercent are on different scales. Winpercent looks especially different because it is a percentage-like variable. 


> Q7. What do you think a zero and one represent for the candy$chocolate column?

For candy$chocolate: 0 = a candy is not chocolate and 1 = a candy is chocolate. 



## Exploratory analysis

> Q8. Plot a histogram of winpercent values using both base R and ggplot2.

```{r}
hist(candy$winpercent)
```

```{r}
library(ggplot2)

ggplot(candy) +
  aes(winpercent) +
  geom_histogram(bins=20)
```

> Q9. Is the distribution of winpercent values symmetrical?

The distribution is not symmetrical. 

> Q10. Is the center of the distribution above or below 50%?

```{r}
mean(candy$winpercent)
median(candy$winpercent)
```

> Q11. On average is chocolate candy higher or lower ranked than fruit candy?

```{r}
mean(candy$winpercent[candy$chocolate == 1])
mean(candy$winpercent[candy$fruity == 1])
```
On average, chocolate candy is higher ranked than fruit candy.

> Q12. Is this difference statistically significant?

```{r}
t.test(
  candy$winpercent[candy$chocolate == 1],
  candy$winpercent[candy$fruity == 1]
)
```

The p-value from the Welch Two Sample t-test is 2.871e-08. Since this is much smaller than 0.05, the difference is statistically significant.



## Overall Candy Rankings

> Q13. What are the five least liked candy types in this set?

```{r}
head(candy[order(candy$sugarpercent, decreasing=TRUE),], n=5)
```

> Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
head(candy[order(candy$sugarpercent),], n=5)
```

> Q15. Make a first barplot of candy ranking based on winpercent values.

```{r, fig.height=15, fig.width=10}
ggplot(candy) + 
  aes(winpercent, rownames(candy)) +
  geom_col()
```

> Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r, fig.height=15, fig.width=10}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col()
```

> Q17. What is the worst ranked chocolate candy?

```{r, fig.height=15, fig.width=10}
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"


ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols) +
  ylab("")
```
The worst ranked chocolate candy is Sixlets. 

> Q18. What is the best ranked fruity candy?

The best ranked fruity candy is Starburst.




## Taking a look at pricepercent


We can fix the label text overplotting with an add on package called `ggrepel` and it's `geom_text_repel`. 
```{r}
# install.packages("ggrepel")
library(ggrepel)

# How about a plot of win vs price
ggplot(candy) +
  aes(x=winpercent, y=pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 5)

```
> Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

```{r}
ord <- order(candy$pricepercent, decreasing = FALSE)
head( candy[ord,c(11,12)], n=5 )
```


> Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?


```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )
```



## Exploring the correlation structure

```{r}
# install.packages("corrplot")
library(corrplot)
cij <- cor(candy)
corrplot(cij)
```
> Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

Fruity and chocolate. 


> Q23. Use your corrplot result to make a prediction: which variables do you expect will have the largest contributions (i.e. loadings) to PC1 (i.e., drive the most separation between candies along the first principal component)?

Fruity and chocolate might have the largest contributions (i.e. loadings) to PC1. 



## Principal Component Analysis

In this case we want to be sure the set `scale=TRUE` because we have one variable `winpercent` that is on a very different scale than all others and would otherwise dominate PCA results. 
```{r}
pca <- prcomp(candy, scale=TRUE)
summary(pca)
```


The first major result figure is the "score plot" of PC1 vs PC2 - how different candy are related to each other on our new PC axis:
```{r}
ggplot(pca$x) +
  aes(PC1, PC2, label=row.names(pca$x)) +
  geom_point(col=my_cols) +
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 5)
```




> Q24. Complete the code to generate the loadings plot above. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you? Where did you see this relationship highlighted previously?

```{r}
ggplot(pca$rotation) +
  aes(PC1, reorder(rownames(pca$rotation), PC1)) +
  geom_col()
```



## Summary


> Q25. Based on your exploratory analysis, correlation findings, and PCA results, what combination of characteristics appears to make a “winning” candy? How do these different analyses (visualization, correlation, PCA) support or complement each other in reaching this conclusion?

A “winning” candy appears to be one that is chocolate-based, often in bar form, and may include richer ingredients like peanut/almond, caramel, nougat, or crisped rice/wafer. It is less likely to be mainly fruity, hard, or sold as pluribus candy. 

The PCA loading plot supports this pattern because chocolate, bar, winpercent, and pricepercent load in the same general direction on PC1, while fruity, hard, and pluribus load in the opposite direction. This suggests that PC1 separates chocolate/bar-style candies from fruity/hard/multiple-piece candies. The correlation plot also supports it as chocolate and fruit are anti-correlated (strong negative correlation coefficients). Other visulization such as ranking of winpercent also indicates that chocolate are more liked. 

