---
title: "HW12"
author: "Zeyu Qi (A17342618)"
format: pdf
---

> Q13: Read this file into R and determine the sample size for each genotype and their corresponding median expression levels for each of these genotypes

```{r}
expr <- read.table("rs8067378_ENSG00000172057.6.txt")
head(expr)
```

```{r}
nrow(expr)
```

```{r}
table(expr$geno)
```

```{r}
aggregate(exp ~ geno, data = expr, FUN = function(x) c(
  sample_size = length(x),
  median_expression = median(x, na.rm = TRUE)
))
aggregate(exp ~ geno, data = expr, FUN = median)
```

Individuals with genotype A/A have a sample size of 108 samples and the highest median expression level: 31.24847

Individuals with genotype A/G have a sample size of 233 samples and an intermediate median expression level: 25.06486

Individuals with genotype G/G have a sample size of 121 samples and the lowest median expression level: 20.07363



> Q14: Generate a boxplot with a box per genotype, what could you infer from the relative expression value between A/A and G/G displayed in this plot? Does the SNP effect the expression of ORMDL3? 

```{r}
library(ggplot2)

ggplot(expr) + aes(geno, exp, fill=geno) +
   geom_boxplot(notch=TRUE)
```

The SNP rs8067378 is associated with ORMDL3 expression, with A/A showing the highest expression and G/G showing the lowest. A allele is associated with higher ORMDL3 expression, while the G allele is associated with lower ORMDL3 expression.







