---
title: "Class 5: Data viz with ggplot2"
author: "Zeyu Qi(A17342618)"
format: pdf
toc: true
---

## Background

There are many graphics systems in R for making plots and figures. These include so-called *"base R" graphics* like the `plot()` function and add on packages like **ggplot2**.

Let's compare how we make a simple figure with these two systems: 

We can use the in-built `cars` dataset:

```{r}
head(cars)
```

```{r}
plot(cars)
```
Before I can use ggplot2 I need to install it on my computer. To do this we can use the function `install.packages("ggplot2")`

> **N.B.** We never run `install.packages()` in our run quarto doc (we run it once only in our R console) as it would re-install the package every time we render. 

Once installed we need to load up the package into our R brain:

```{r}
library(ggplot2)
```



The main function in the **ggplot2** packages is called `ggplot()`
```{r}
ggplot(cars)
```

Every ggplot has at least 3 layers: 

- the **data** (a datafrome)
- the **aes**thetics (how data maps to plot)
- the **geom** layer (how you want the plot drawn)

```{r}
ggplot(cars) + 
  aes(x=speed, y=dist) +
  geom_point()
```

## Add some custom features

Let's add a trend line taht shows the relationship between speed and distance. 

```{r}
ggplot(cars) + 
  aes(x=speed, y=dist) +
  geom_point() + 
  geom_smooth() +
  theme_bw() +
  labs(title="Stoping Dist of old cars", 
       x="Speed (MPH)",
       y="Distance (ft)")
```

> Q. Can you make the `geom_smoothe()` function produce a linear straight line fit to the data and turn off grey region. 

---
## Gene Expression

```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)
```


```{r}
sum(genes$State == "up")
```

A useful new function in this context is the `table()` function:

```{r}
table(genes$State)
```
```{r}
ggplot(genes) +
  aes(Condition1, Condition2, col=State) +
  geom_point() + 
  scale_color_manual(values=c("red", "gray", "blue")) +
  theme_bw() +
  labs(x="No Drug", y="Drug", title="Expresion changes upon GLP-1 inhibitor treatment")
```


## Going Further

Here we read the famous gapmider dataset: 

```{r}
# File location online
url <- "https://raw.githubusercontent.com/jennybc/gapminder/master/inst/extdata/gapminder.tsv"

gapminder <- read.delim(url)
head(gapminder)
```


> Q How many entries (i.e rows) are in this dataset?


```{r}
nrow(gapminder)
```

> Q. How many different "country" entries are in this dataset?

```{r}
length(table(gapminder$country))
```

```{r}
length(unique(gapminder$country))
```

Let's make our first plot of the entire dataset:

```{r}
p <- ggplot(gapminder) + 
  aes(gdpPercap, lifeExp, col=continent) +
  geom_point(alpha=0.3) 
```

```{r}
p
```


I can add more layers to `p`
```{r}
p + 
  facet_wrap(~continent)
```


Make a plot for 1977 and 2007 only (not all the years in the dataset).

> Q. First use the **dplyr** package and the `filter()` function from that pacakge to extract the year 2007.

```{r, message=FALSE}
library(dplyr)
```


```{r}
ggplot(filter(gapminder, year == 2007 | year == 1977)) + 
  aes(gdpPercap, lifeExp, col=continent, size=pop) +
  geom_point() + 
  facet_wrap(~year)
```

> Q. Make a histogram of lifeExp faceted by continent

```{r}
ggplot(gapminder) +
  aes(lifeExp) +
  geom_histogram() + 
  facet_wrap(~continent)
```


> Q. Make a histogram of lifeExp colored by continent

```{r}
ggplot(gapminder) +
  aes(lifeExp, fill=continent) +
  geom_histogram()
```